### 2015-04-28 thesis-proposal/parsing.tex
Updated proposal
Updated poropsal

In this work, we will describe an alternative parsing strategy that avoids these problems by shifting responsibility for parsing certain \emph{generic literal forms} into the typechecker. The typechecker, in turn, defers responsibility to user-defined types, by treating the body of the literal as a term of the   \emph{type- or type-constructor-specific language (TSL)} associated with the type it is being checked against. The TSL is responsible for rewriting this term to ultimately use only general-purpose notation. This strategy avoids the problem of conflicting syntax, because neither the base language nor TSLs are ever extended directly. It also permits semantic flexibility -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, maps and other data structures, like JSON literals. This frees these common notations from being tied to the variant of a  data structure built into a language's standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
\begin{figure}[t]
\begin{lstlisting}
let imageBase : URL = <SURLimages.example.comEURL>
let bgImage : URL = <SURL%EURLimageBaseSURL%/background.pngEURL>
new : SearchServer
  def resultsFor(searchQuery, page)
    serve(~) (* serve : HTML -> Unit *)
SHTML      >html
        >head
          >title Search Results
          >style EHTML~
SCSS            body { background-image: url(%ECSSbgImageSCSS%) }
            #search { background-color: %ECSSdarken(`SCOLOR#aabbccECOLOR`, SPCT10pctEPCT)SCSS% }
ECSSSHTML        >body
          >h1 Results for < EHTMLHTML.Text(searchQuery)SHTML
          >div[id="search"]
            Search again: < EHTMLSearchBox($\texttt{"}$SSTRGo!ESTR$\texttt{"}$)SHTML
          <EHTML (* fmt_results : DB * SQLQuery * Nat * Nat -> HTML *)
             fmt_results(db, ~, SNAT10ENDNAT, page)
               SSQLSELECT * FROM products WHERE {ESQLsearchQuerySSQL} in titleESQL
\end{lstlisting}
%\vspace{-8px}
\caption{Wyvern Example with Multiple TSLs}
\label{f-example}
%\vspace{-10px}
\end{figure}
%\begin{lstlisting}
%let empty_set : Set = { }
%let empty_dict : Dict = { }
%let empty_json : JSON = { }
%\end{lstlisting}
\subsection{Wyvern}
We specify our work as a variant of a new programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}. To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects other than non-termination) and it does not enforce a uniform access principle for objects (fields can be accessed directly). Objects can thus be thought of as recursive labeled products with support for simple methods (functions that are automatically given a self-reference) for convenience. We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML. One can refer to the version of the language described in this thesis as \emph{TSL Wyvern}. TSL Wyvern has a layout-sensitive concrete syntax, for reasons we will discuss.

@@

In addition to supporting conventional notation for URLs, this TSL supports \emph{splicing}
another Wyvern expression of type \li{URL} to form a larger URL. The spliced term is delimited by percent signs,
as seen on line 2 of Fig. \ref{f-example}. The TSL parses code between percent signs  as a Wyvern expression, using its abstract syntax tree (AST) to construct an AST for the expression as a whole. A string-based representation of the URL is never constructed. Note that the delimiters used to go from Wyvern to a TSL are controlled by Wyvern  while the TSL controls how to return to Wyvern. 
%\subsection{General-Purpose Notation for Object and Case Types}
%The general-purpose introductory form for object types is \li{new}. This form is a syntactic \emph{forward reference} to the layout-delimited block of {definitions} beginning on the line immediately after the line \li{new} appears in (line 4 in this case), and ending when the indentation level has returned to the baseline, or when the file ends (after line 19 in this case). An object in TSL Wyvern can contain methods, introduced using \li{def}, and fields, introduced using \li{val}. Here we have just a single method, \li{serve_results} taking two arguments. Object types in TSL Wyvern are simple structural interfaces that constrain the signatures of fields and methods. The \emph{type ascription} around \li{new} checks that the object being introduced satisfies the signature of \li{SearchServer} (not shown).\todo{discuss case types}
\subsection{Layout-Delimited Literals}

On line 5 of Fig. \ref{f-example}, we see a call to a function \li{serve} (not shown) which has type \li{HTML -> Unit}. Here, \li{HTML} is a user-defined \emph{case type}, having cases for each HTML tag as well as some other structures, like text nodes and sequencing. Declarations  of some of these cases can be seen on lines 2-6 of Fig. \ref{f-htmltype} (note that TSL Wyvern also includes simple product types for convenience, written \li{T1 * T2}). We could again use Wyvern's general-purpose introductory form for case types, e.g. \li{BodyElement((attrs, child))}. But, as discussed above, using this syntax can be inconvenient and  cognitively demanding. Thus, we associate a TSL with \li{HTML} that provides a simplified notation for writing HTML, shown being used on lines 6-18 of Fig. \ref{f-example}. This literal body is layout-delimited, rather than delimited by explicit tokens as in Fig. \ref{f-delims}, and introduced by a form of \emph{forward reference}, written \li{~} (``tilde''), on the previous line. Because the forward reference occurs in a position where the expected type is \li{HTML}, the literal body is governed by that type's TSL. The forward reference will be replaced by the general-purpose term, of type \li{HTML}, generated by the TSL during typechecking. Because layout was used as a delimiter, there are no syntactic constraints on the body, unlike with inline literals (Fig. \ref{f-delims}). For HTML, this is quite useful, as all of the inline forms impose constrains that would cause conflict with some valid HTML.
\subsection{Implementing a TSL}
\begin{figure}
\begin{lstlisting}[escapechar=$]
casetype HTML 
  Empty
  Seq of HTML * HTML 
  Text of String
  BodyElement of Attributes * HTML
  StyleElement of Attributes * CSS
  (* ... *)
  metadata : HasTSL = new
    val parser = ~
SGRM      start <- '>body'= attributes> start>
        EGRMfn attrs, child => `SQTHTML.BodyElement((%EQTattrsSQT%, %EQTchildSQT%))EQT`
SGRM      start <- '>style'= attributes> EXP>
        EGRMfn attrs, e => `SQTHTML.StyleElement((%EQTattrsSQT%, %EQTeSQT%))EQT`
SGRM      start <- '<'= EXP>EGRM 
        fn e => `SQT%EQTeSQT%EQT`
\end{lstlisting}
\vspace{-8px}
\caption{A Wyvern case type with an associated TSL.}
\label{f-htmltype}
\end{figure}
\begin{figure}[t]
\begin{subfigure}[t]{.58\textwidth}
\begin{lstlisting}
objtype HasTSL
  val parser : Parser
objtype Parser                          
  def parse(ps : ParseStream) : ParseResult
  metadata : HasTSL = new
    val parser = (* parser generator *)
casetype ParseResult
  OK of Exp * ParseStream
  Error of String * Location  
\end{lstlisting}
\end{subfigure}
\begin{subfigure}[t]{.42\textwidth}
\begin{lstlisting}[linewidth=.42\textwidth,firstnumber=10]
casetype Exp 
  Var of ID
  Lam of ID * Type * Exp
  Ap of Exp * Exp
  Tuple of Exp * Exp
  ... 
  Spliced of ParseStream
  metadata : HasTSL = new
    val parser = (* quasiquotes *)
\end{lstlisting}
\end{subfigure}
\caption{Some of the types included in the Wyvern prelude. They are mutually defined.}
%\vspace{-10px}
\label{f-builtins}
%\label{fig:typeParser}
\end{figure}
Portions of the implementation of the TSL for \li{HTML} are shown on lines 8-15 of Fig. \ref{f-htmltype}. A TSL is associated with a named type, forming an \emph{active type}, using a more general mechanism for associating a pure, static value with a named type, called its \emph{metadata}. Metadata is introduced as shown on line 8 of Fig. \ref{f-htmltype}. Type metadata, in this context, is comparable to class annotations in Java or attributes in C\#/F\# and internalizes the practice of writing metadata using comments, so that it can be checked by the language and accessed programmatically more easily. This can be used for a variety of purposes -- to associate documentation with a type, to mark types as being deprecated, and so on.

For the purposes of this work, metadata values will always be of type \li{HasTSL}, an object type that declares a single field, \li{parser}, of type \li{Parser}. The \li{Parser} type is an object type declaring a single method, \li{parse}, that transforms a \li{ParseStream} extracted from a literal body to a Wyvern AST. An AST is a value of type \li{Exp}, a case type that encodes the abstract syntax of Wyvern expressions. Fig. \ref{f-builtins} shows portions of the declarations of these types, which live in the Wyvern \emph{prelude} (a collection of types that are automatically loaded before any other).

@@

\subsection{Remaining Tasks and Timeline}
This work was published at ECOOP 2014 (and received a Distinguished Paper Award) \cite{TSLs}. We have not formally shown how parameterized types work (e.g. we can't yet implement list literal syntax uniformly as a library). I plan to do this as remaining work in the course of completing the dissertation writing. I am currently working with two undergraduates on other extensions to this work, but do not plan to include it in my dissertation. % and we will defer these tasks to after the OOPSLA deadline in late March.% If not accepted, we will aim to resubmit to OOPSLA.
%\begin{enumerate}
%\item In our current top-level grammar, expressions within parenthesis must still obey layout constraints. This is not strictly necessary for preventing ambiguities, and it is quite convenient to not require this (as in Python, for example). We will add these simple rules to the grammar.
%\item 
%\item In macro systems like Scheme's, free variables in generated ASTs refer to their bindings within the macro definition, rather than where they are inserted. To support this, we have introduced the \li{toast} operator. Using it is currently explicit, so free variables that were not inside \li{toast} result in errors. We believe we can insert \li{toast} automatically using single variable (rather than whole expression) splicing so that free variables are implicitly \li{toast}ed, from within a library. We will show this.
%\item To examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code. We need to finish this (collaborative work with Darya Kurilova).
%\end{enumerate}
%\item The corpus analysis we conducted was preliminary. We must perform this in a more complete and rigorous manner (collaborative work with Darya Kurilova).
%\item The current implementation of Wyvern does not include all of these mechanisms as described. Although the implementation of Wyvern is not a project I am responsible for, I plan on working with the student who is doing this to implement as much of what we have described as possible.\todo{should I drop this from the proposal? I don't want to promise that everything will be implemented because that is largely contingent on Benjamin's effort, not mine.}
%\end{enumerate}
%
%\subsection{TODO}
%- Figure out separate compilation (e.g. of CSS)
\newpage

>>>>

In this work, we will describe an alternative parsing strategy that avoids these problems by shifting responsibility for parsing certain \emph{generic literal forms} into the typechecker. The typechecker, in turn, defers responsibility to user-defined types, by treating the body of the literal as a term of the   \emph{type- or type-constructor-specific language (TSL)} associated with the type it is being checked against. The TSL is responsible for rewriting this term to ultimately use only general-purpose notation. This strategy avoids the problem of conflicting syntax, because neither the base language nor TSLs are ever extended directly. It also permits semantic flexibility -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, maps and other data structures, like JSON literals. This frees these common notations from being tied to the variant of a  data structure built into a language's standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
\begin{figure}[t]
\begin{lstlisting}
let imageBase : URL = <SURLimages.example.comEURL>
let bgImage : URL = <SURL%EURLimageBaseSURL%/background.pngEURL>
new : SearchServer
  def resultsFor(searchQuery, page)
    serve(~) (* serve : HTML -> Unit *)
SHTML      >html
        >head
          >title Search Results
          >style EHTML~
SCSS            body { background-image: url(%ECSSbgImageSCSS%) }
            #search { background-color: %ECSSdarken(`SCOLOR#aabbccECOLOR`, SPCT10pctEPCT)SCSS% }
ECSSSHTML        >body
          >h1 Results for < EHTMLHTML.Text(searchQuery)SHTML
          >div[id="search"]
            Search again: < EHTMLSearchBox($\texttt{"}$SSTRGo!ESTR$\texttt{"}$)SHTML
          <EHTML (* fmt_results : DB * SQLQuery * Nat * Nat -> HTML *)
             fmt_results(db, ~, SNAT10ENDNAT, page)
               SSQLSELECT * FROM products WHERE {ESQLsearchQuerySSQL} in titleESQL
\end{lstlisting}
%\vspace{-8px}
\caption{Wyvern Example with Multiple TSLs}
\label{f-example}
%\vspace{-10px}
\end{figure}
%\begin{lstlisting}
%let empty_set : Set = { }
%let empty_dict : Dict = { }
%let empty_json : JSON = { }
%\end{lstlisting}
\subsection{Wyvern}
We specify our work as a variant of a new programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}. To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects other than non-termination) and it does not enforce a uniform access principle for objects (fields can be accessed directly). Objects can thus be thought of as recursive labeled products with support for simple methods (functions that are automatically given a self-reference) for convenience. We also add recursive labeled sum types, which we call \emph{variant types}, that are quite similar to datatypes in ML. One can refer to the version of the language described in this thesis as \emph{TSL Wyvern}. TSL Wyvern has a layout-sensitive concrete syntax, for reasons we will discuss.

@@

In addition to supporting conventional notation for URLs, this TSL supports \emph{splicing}
another Wyvern expression of type \li{URL} to form a larger URL. The spliced term is delimited by percent signs,
as seen on line 2 of Fig. \ref{f-example}. The TSL parses code between percent signs  as a Wyvern expression, using its abstract syntax tree (AST) to construct an AST for the expression as a whole. A string-based representation of the URL is never constructed. Note that the delimiters used to go from Wyvern to a TSL are controlled by Wyvern  while the TSL controls how to return to Wyvern. 
%\subsection{General-Purpose Notation for Object and variant types}
%The general-purpose introductory form for object types is \li{new}. This form is a syntactic \emph{forward reference} to the layout-delimited block of {definitions} beginning on the line immediately after the line \li{new} appears in (line 4 in this case), and ending when the indentation level has returned to the baseline, or when the file ends (after line 19 in this case). An object in TSL Wyvern can contain methods, introduced using \li{def}, and fields, introduced using \li{val}. Here we have just a single method, \li{serve_results} taking two arguments. Object types in TSL Wyvern are simple structural interfaces that constrain the signatures of fields and methods. The \emph{type ascription} around \li{new} checks that the object being introduced satisfies the signature of \li{SearchServer} (not shown).\todo{discuss variant types}
\subsection{Layout-Delimited Literals}

On line 5 of Fig. \ref{f-example}, we see a call to a function \li{serve} (not shown) which has type \li{HTML -> Unit}. Here, \li{HTML} is a user-defined \emph{variant type}, having cases for each HTML tag as well as some other structures, like text nodes and sequencing. Declarations  of some of these cases can be seen on lines 2-6 of Fig. \ref{f-htmltype} (note that TSL Wyvern also includes simple product types for convenience, written \li{T1 * T2}). We could again use Wyvern's general-purpose introductory form for variant types, e.g. \li{BodyElement((attrs, child))}. But, as discussed above, using this syntax can be inconvenient and  cognitively demanding. Thus, we associate a TSL with \li{HTML} that provides a simplified notation for writing HTML, shown being used on lines 6-18 of Fig. \ref{f-example}. This literal body is layout-delimited, rather than delimited by explicit tokens as in Fig. \ref{f-delims}, and introduced by a form of \emph{forward reference}, written \li{~} (``tilde''), on the previous line. Because the forward reference occurs in a position where the expected type is \li{HTML}, the literal body is governed by that type's TSL. The forward reference will be replaced by the general-purpose term, of type \li{HTML}, generated by the TSL during typechecking. Because layout was used as a delimiter, there are no syntactic constraints on the body, unlike with inline literals (Fig. \ref{f-delims}). For HTML, this is quite useful, as all of the inline forms impose constrains that would cause conflict with some valid HTML.
\subsection{Implementing a TSL}
\begin{figure}
\begin{lstlisting}[escapechar=$]
variant HTML 
  Empty
  Seq of HTML * HTML 
  Text of String
  BodyElement of Attributes * HTML
  StyleElement of Attributes * CSS
  (* ... *)
  metadata : HasTSL = new
    val parser = ~
SGRM      start <- '>body'= attributes> start>
        EGRMfn attrs, child => `SQTHTML.BodyElement((%EQTattrsSQT%, %EQTchildSQT%))EQT`
SGRM      start <- '>style'= attributes> EXP>
        EGRMfn attrs, e => `SQTHTML.StyleElement((%EQTattrsSQT%, %EQTeSQT%))EQT`
SGRM      start <- '<'= EXP>EGRM 
        fn e => `SQT%EQTeSQT%EQT`
\end{lstlisting}
\vspace{-8px}
\caption{A Wyvern variant type with an associated TSL.}
\label{f-htmltype}
\end{figure}
\begin{figure}[t]
\begin{subfigure}[t]{.58\textwidth}
\begin{lstlisting}
objtype HasTSL
  val parser : Parser
objtype Parser                          
  def parse(ps : ParseStream) : ParseResult
  metadata : HasTSL = new
    val parser = (* parser generator *)
variant ParseResult
  OK of Exp * ParseStream
  Error of String * Location  
\end{lstlisting}
\end{subfigure}
\begin{subfigure}[t]{.42\textwidth}
\begin{lstlisting}[linewidth=.42\textwidth,firstnumber=10]
variant Exp 
  Var of ID
  Lam of ID * Type * Exp
  Ap of Exp * Exp
  Tuple of Exp * Exp
  ... 
  Spliced of ParseStream
  metadata : HasTSL = new
    val parser = (* quasiquotes *)
\end{lstlisting}
\end{subfigure}
\caption{Some of the types included in the Wyvern prelude. They are mutually defined.}
%\vspace{-10px}
\label{f-builtins}
%\label{fig:typeParser}
\end{figure}
Portions of the implementation of the TSL for \li{HTML} are shown on lines 8-15 of Fig. \ref{f-htmltype}. A TSL is associated with a named type, forming an \emph{active type}, using a more general mechanism for associating a pure, static value with a named type, called its \emph{metadata}. Metadata is introduced as shown on line 8 of Fig. \ref{f-htmltype}. Type metadata, in this context, is comparable to class annotations in Java or attributes in C\#/F\# and internalizes the practice of writing metadata using comments, so that it can be checked by the language and accessed programmatically more easily. This can be used for a variety of purposes -- to associate documentation with a type, to mark types as being deprecated, and so on.

For the purposes of this work, metadata values will always be of type \li{HasTSL}, an object type that declares a single field, \li{parser}, of type \li{Parser}. The \li{Parser} type is an object type declaring a single method, \li{parse}, that transforms a \li{ParseStream} extracted from a literal body to a Wyvern AST. An AST is a value of type \li{Exp}, a variant type that encodes the abstract syntax of Wyvern expressions. Fig. \ref{f-builtins} shows portions of the declarations of these types, which live in the Wyvern \emph{prelude} (a collection of types that are automatically loaded before any other).

@@

\subsection{Remaining Tasks and Timeline}
This work was published at ECOOP 2014 (and received a Distinguished Paper Award) \cite{TSLs}. We have not formally shown how parameterized and abstract types work (e.g. we can't yet implement list literal syntax uniformly as a library). I plan to do this as remaining work in the course of completing the dissertation writing. I am currently working with two undergraduates on other extensions to this work, but do not plan to include it in my dissertation. % and we will defer these tasks to after the OOPSLA deadline in late March.% If not accepted, we will aim to resubmit to OOPSLA.
%\begin{enumerate}
%\item In our current top-level grammar, expressions within parenthesis must still obey layout constraints. This is not strictly necessary for preventing ambiguities, and it is quite convenient to not require this (as in Python, for example). We will add these simple rules to the grammar.
%\item 
%\item In macro systems like Scheme's, free variables in generated ASTs refer to their bindings within the macro definition, rather than where they are inserted. To support this, we have introduced the \li{toast} operator. Using it is currently explicit, so free variables that were not inside \li{toast} result in errors. We believe we can insert \li{toast} automatically using single variable (rather than whole expression) splicing so that free variables are implicitly \li{toast}ed, from within a library. We will show this.
%\item To examine how broadly applicable the technique is, we conduct a simple corpus analysis, finding that string languages are used ubiquitously in existing Java code. We need to finish this (collaborative work with Darya Kurilova).
%\end{enumerate}
%\item The corpus analysis we conducted was preliminary. We must perform this in a more complete and rigorous manner (collaborative work with Darya Kurilova).
%\item The current implementation of Wyvern does not include all of these mechanisms as described. Although the implementation of Wyvern is not a project I am responsible for, I plan on working with the student who is doing this to implement as much of what we have described as possible.\todo{should I drop this from the proposal? I don't want to promise that everything will be implemented because that is largely contingent on Benjamin's effort, not mine.}
%\end{enumerate}
%
%\subsection{TODO}
%- Figure out separate compilation (e.g. of CSS)
\newpage
